---
title: "R Notebook"
output: html_notebook
---

Initialization of Data and packages
The Dataframe is a little messy atm, response variables are a little awkward as we are now just numbering different bone details, can probably build up a new data-frame from this later if we wanna do within bone stuff like for covariances


```{r}
#Make sure to set wd depending on location, H:/ on laptop, Z:/ on desktop
library("mice")
library("ape")
Mars <- read.csv("mars.csv", header=TRUE)
Mars
Mars_Avg <- read.csv("mars_avg.csv")
Mars_Avg
Marsimp <- Mars

as.factor(Marsimp$Species)

# Need to do some work on the df to input phylogenetic information
```

Run an initial Mice run with 5 iterations

```{r}

# Initialize mice current example is 
imp2 <- mice(Marsimp, maxit=5, print = FALSE, seed = 123)

stripplot(imp2, dentary ~.imp)

# Extract complete imputed data

imp2.dat <- complete(imp2, action = "long")

# Collect Mean values and Variance on the Mean of all variables
MEAN <- tapply(imp2.dat$dentary, imp2.dat$.imp, mean)
VAR.MEAN <- tapply(imp2.dat$dentary, imp2.dat$.imp, var)/nrow(imp2$data)

cbind("MEAN" = MEAN, "VAR.MEAN" = VAR.MEAN)

#Plot Mean and Variance of Mean for an example variable, in this case Articular Width
par(mfrow = c(1,2))
plot(MEAN ~ I(1:imp2$m), ylab= "Mean", xlab = "Imputation", main = "Mean \nby imputation")
abline(h=mean(MEAN), col = "red", lty = 2, lwd=2)

plot(VAR.MEAN~I(1:imp2$m), ylab= "Variance of Mean", xlab = "Imputation", main = "Variance of Mean\nby imputation")
abline(h=mean(VAR.MEAN), col = "red", lty = 2, lwd=2)

# Pool the means and variance of the means
POOLED <- pool.scalar(MEAN, VAR.MEAN, nrow(imp2$data))

data.frame(pooled.mean = mean(MEAN), mean.var = mean(VAR.MEAN), btw.var = var(MEAN), pooled.var = mean(VAR.MEAN) + (1+1/length(MEAN))*var(MEAN))


```
Obtained Tree from https://timetree.org/, also awaiting some from Vertlife.org, but still need to receive them

Most of the stuff in this chunk is just Tree iteration work, potentially useful later but not so much right now

```{r}


mars_tree <- read.tree("newtree.nwk")
plot(mars_tree)
Mars_Avg

mars_tree2 <- ladderize(mars_tree)
mars_avg2 <- Mars_Avg[which(Mars_Avg$X %in% mars_tree$tip.label),]
rownames(mars_avg2) <- mars_avg2$X
mars_avg2 <- mars_avg2[match(mars_tree$tip.label, rownames(mars_avg2)),]

```

```{r}
#Set a new column in  dataframe denoting the node value of the tip of the organism
for (i in 1:length(mars_tree$tip.label)) {
  for (j in 1:length(mars_avg2$X)) {
    if (mars_tree$tip.label[i] == mars_avg2$X[j]) {
      mars_avg2$tip[j] <- i
    }
  }
}

nodepath(mars_tree)
node.depth.edgelength(mars_tree)


timepath <- nodepath(mars_tree)


#Make this a function, very useful actually

nodes <- c()
timepath <- list()
iter <- 1
for (i in nodepath(mars_tree)) {
  vec <- c()
  for (j in i) {
    # if (j %in% nodes) {
    #  next
    #}
    vec <- c(vec,node.depth.edgelength(mars_tree)[j])
    #nodes <- c(nodes, j)
  }
  timepath[[iter]] <- vec
  iter = iter + 1
}

#now i can obtain tip values from Mars_Avg, and I have time scaling so I can evolve the tree towards or away from those values

#have to figure out how to deal with path redundancy

#Tip Paths - required for all - nothing to say that identical paths in the nodes are redundant so this is kinda useless
tippaths <- list()
for (i in 1:length(timepath)) {
  k <- length(timepath[[i]])
  tipvec <- c()
  print(k)
  for (j in 1:k) {
    #print(tipvec)
    if (j >= k-1) {
      print(j)
      tipvec <- c(tipvec, timepath[[i]][j])
      print(tipvec)
    }  
  }
  tippaths[[i]] <- tipvec
}
tippaths

#same for nodes? Then we can do BM for each tip path from prior results

```

Adding the initial averaged data to the tree, and performing Independent Contrasts and Phylogenetic Eigenvector Regression


```{r}

library(phytools)
#Data fixing so tree matches given data
mars_dat <- mars_avg2[,c("dentary", "dentary.1", "dentary.2")]
mars_dat


# Independent Contrasts tests
x1 <- pic(mars_avg2$dentary[-21], mars_tree2, scaled = FALSE)
x2 <- pic(mars_avg2$dentary.1[-21], mars_tree2, scaled = FALSE)

z <- lm(x1~x2-1)

#Phylogenetic Eigenvector regression on initial Data
x <- PVRdecomp(mars_tree2, scale = TRUE)
trait2 <- mars_dat2$dentary
y = PVR(x, trait = trait2)

```

Doing the same with imputed data

```{r}

library("PVR")
library("plyr")
MEANS <- ddply(imp2.dat, .(Species), summarize, mean=mean(dentary))


#Need to match tree and complete data (cause tree is missing some species)
MEANS2 <- MEANS[which(MEANS$Species %in% mars_tree2$tip.label),]
missingfromtree <- which(!(mars_tree2$tip.label %in% MEANS$Species))
mars_tree3 <- drop.tip(mars_tree2, missingfromtree)
rownames(MEANS2) <- MEANS2$Species
MEANS2 <- MEANS2[match(mars_tree2$tip.label, rownames(MEANS2)),]
trait <- MEANS2[,2]
X <- PVRdecomp(mars_tree, scale = TRUE)
#y2 = PVR(X, trait = trait, method = "moran")


```

Now we need to Figure out how to get the phylogeny to influence our Missing Data Imputation

```{r}
missingfromtree2 = which(!(Marsimp$Species %in% mars_tree$tip.label))

marsimp2 = Marsimp[-missingfromtree2,]
marsimp2$Species = as.factor(marsimp2$Species)
marsimp2
library("PVR")
library("plyr")
X <- PVRdecomp(mars_tree, scale = FALSE)

evec <- X@Eigen$vectors

edf <- data.frame(evec)
edf$Species <- mars_tree$tip.label

marsimp3 <- merge(marsimp2, edf, by="Species")

imp0 <- mice(marsimp2, maxit = 0)
predM <- imp0$predictorMatrix
impM <- imp0$method
impM[c(3:25, 27:29)] <- "norm"


traits = colnames(Marsimp)[-1]
traits = traits[-1]

predM1 <- predM
predM1[traits, "Species"] <- as.integer(-2)
predM1["Individual", "Species"] <- as.integer(-2)
impM1 <- impM
impM1["Species"] <- c("2lonly.norm")
impM1[c(3:25, 27:29)] <- c("2l.norm")
marsimp2$Species <- as.factor(marsimp2$Species)
Species_List = marsimp3$Species
marsimp2$Species <- sapply(marsimp2$Species, unclass)

predM4 <- predM1
impM4 <- impM1
impM4["Species"] <- c("2lonly.pmm")
impM4[c(3:25, 27:29)] <- c("2l.pmm")

impM5 <- impM1
impM5["Species"] <- c("2lonly.pmm")
impM5[c(3:25, 27:29)] <- c("2l.pan")


impM3 <- impM
impM3[c(3:25, 27:29)] <- c("norm")

imp1 <- mice(marsimp3, maxit = 0)
predMpvr <- imp1$predictorMatrix
impMpvr <- imp1$method
impMpvr[c(3:25, 27:29)] <- "norm"


traits = colnames(Marsimp)[-1]
traits = traits[-1]

predM2 <- predMpvr
predM2[traits, "Species"] <- as.integer(-2)
predM2["Species", c(3:25, 27:29, 31:62)] <- as.integer(-2)
predM2["Individual", "Species"] <- as.integer(-2)
impM2 <- impMpvr
impM2["Species"] <- c("2lonly.pmm")
impM2[c(3:25, 27:29)] <- c("2l.pan")
impM2[c(31:62)] <- c("")
marsimp3$Species <- as.factor(marsimp3$Species)
Species_List = marsimp3$Species
marsimp3$Species <- sapply(marsimp3$Species, unclass)
marsimp3$Species <- as.integer(marsimp3$Species)

```

```{R}

impcart <- mice(marsimp3, maxit = 20, method = "cart")
impmidas <- mice(marsimp2, maxit = 20, method = "midastouch")
impnorm <- mice(marsimp3, maxit = 20, predictorMatrix =  predM, method = impM3)
imp4 <- mice(marsimp2, maxit = 20, predictorMatrix = predM1, method = impM1)

library("miceadds")
impnopvr <- mice(marsimp2, maxit = 5, predictorMatrix = predM1, method = impM1)
impnopvrpmm <- mice(marsimp2, maxit = 20, predictorMatrix = predM1, method = impM4)

imppvr2 <- mice(marsimp3, m = 5, maxit = 20, predictorMatrix = predM2, method = impM2)
#imp3 <- mice(marsimp3, m = 25, maxit=25, print = FALSE, seed = 123)
densityplot(imp4, ~femur)
stripplot(imp4, dentary ~.imp)
plot(imp4)
# Extract complete imputed data

imp3.dat <- complete(imp4, action = "long")

newdata = split(imp3.dat, f = imp3.dat$.imp)
Output = data.frame(Species = sort(edf$Species))
Filename = "MI_Data2lonly/MarsMI_Imp"
iter = 0
variables = colnames(Marsimp)

```

```{R}

impcart <- mice(marsimp3, maxit = 20, method = "cart")
impmmpvr <- mice(marsimp3, maxit = 20, predictorMatrix = predM2)


impmidas <- mice(marsimp2, maxit = 20, predictorMatrix = predM1, method = "midastouch")
impnopvr <- mice(marsimp2, maxit = 5, predictorMatrix = predM1, method = impM1)
impnopvrpmm <- mice(marsimp2, maxit = 20, predictorMatrix = predM1, method = impM4)
impnopvrpmm <- mice(marsimp2, maxit = 20, predictorMatrix = predM1, method = impM5)







for (i in newdata) {
  current = i
  current$Species = Species_List
  var_idx = which(colnames(i) %in% variables)
  iter = iter + 1
  Output = current[,var_idx]
  newfile = paste(Filename, iter, sep = "_")
  newfile = paste(newfile, "csv", sep = ".")
  write.csv(Output, newfile, row.names = FALSE)
}


```

```{r}


newdata = split(imp3.dat, f = imp3.dat$.imp)
Output = data.frame(Species = sort(edf$Species))
iter = 1
for (i in newdata) {
  current = i
  iter = iter + 1
  MEANS <- ddply(i, .(Species), summarize, mean=mean(dentary))
  Output[,iter] = MEANS$mean
}
Filename = "Data/MarsMI_Dentary_ArticularWidth.csv"
write.csv(Output, Filename, row.names = FALSE)

Output = data.frame(Species = sort(edf$Species))
iter = 1
for (i in newdata) {
  current = i
  iter = iter + 1
  MEANS <- ddply(i, .(Species), summarize, mean=mean(dentary.1))
  Output[,iter] = MEANS$mean
}
Filename = "Data/MarsMI_Dentary_ArticularLength.csv"
write.csv(Output, Filename, row.names = FALSE)

Output = data.frame(Species = sort(edf$Species))
iter = 1
for (i in newdata) {
  current = i
  iter = iter + 1
  MEANS <- ddply(i, .(Species), summarize, mean=mean(dentary.2))
  Output[,iter] = MEANS$mean
}
Filename = "Data/MarsMI_Dentary_ForamenLength.csv"
write.csv(Output, Filename, row.names = FALSE)

Output = data.frame(Species = sort(edf$Species))
iter = 1
for (i in newdata) {
  current = i
  iter = iter + 1
  MEANS <- ddply(i, .(Species), summarize, mean=mean(dentary.3))
  Output[,iter] = MEANS$mean
}
Filename = "Data/MarsMI_Dentary_RowWidthatm4.csv"
write.csv(Output, Filename, row.names = FALSE)

Output = data.frame(Species = sort(edf$Species))
iter = 1
for (i in newdata) {
  current = i
  iter = iter + 1
  MEANS <- ddply(i, .(Species), summarize, mean=mean(dentary.4))
  Output[,iter] = MEANS$mean
}
Filename = "Data/MarsMI_Dentary_Heightatm4.csv"
write.csv(Output, Filename, row.names = FALSE)

Output = data.frame(Species = sort(edf$Species))
iter = 1
for (i in newdata) {
  current = i
  iter = iter + 1
  MEANS <- ddply(i, .(Species), summarize, mean=mean(dentary.5))
  Output[,iter] = MEANS$mean
}
Filename = "Data/MarsM_Dentary_Heightatp3.csv"
write.csv(Output, Filename, row.names = FALSE)

Output = data.frame(Species = sort(edf$Species))
iter = 1
for (i in newdata) {
  current = i
  iter = iter + 1
  MEANS <- ddply(i, .(Species), summarize, mean=mean(scapula))
  Output[,iter] = MEANS$mean
}
Filename = "Data/MarsMI_scapula_infraspinouswidth.csv"
write.csv(Output, Filename, row.names = FALSE)

Output = data.frame(Species = sort(edf$Species))
iter = 1
for (i in newdata) {
  current = i
  iter = iter + 1
  MEANS <- ddply(i, .(Species), summarize, mean=mean(scapula.1))
  Output[,iter] = MEANS$mean
}
Filename = "Data/MarsMI_scapula_infraspinousLength.csv"
write.csv(Output, Filename, row.names = FALSE)

Output = data.frame(Species = sort(edf$Species))
iter = 1
for (i in newdata) {
  current = i
  iter = iter + 1
  MEANS <- ddply(i, .(Species), summarize, mean=mean(scapula.2))
  Output[,iter] = MEANS$mean
}
Filename = "Data/MarsMI_Scapula_GlenoidArticularFacetLength.csv"
write.csv(Output, Filename, row.names = FALSE)

Output = data.frame(Species = sort(edf$Species))
iter = 1
for (i in newdata) {
  current = i
  iter = iter + 1
  MEANS <- ddply(i, .(Species), summarize, mean=mean(scapula.3))
  Output[,iter] = MEANS$mean
}
Filename = "Data/MarsMI_Scapula_GlenoidMaxLength.csv"
write.csv(Output, Filename, row.names = FALSE)

Output = data.frame(Species = sort(edf$Species))
iter = 1
for (i in newdata) {
  current = i
  iter = iter + 1
  MEANS <- ddply(i, .(Species), summarize, mean=mean(scapula.4))
  Output[,iter] = MEANS$mean
}
Filename = "Data/MarsMI_Scapula_GlenoidWidth.csv"
write.csv(Output, Filename, row.names = FALSE)

Output = data.frame(Species = sort(edf$Species))
iter = 1
for (i in newdata) {
  current = i
  iter = iter + 1
  MEANS <- ddply(i, .(Species), summarize, mean=mean(scapula.5))
  Output[,iter] = MEANS$mean
}
Filename = "Data/MarsMI_Scapula_AngleScapulaNeck.csv"
write.csv(Output, Filename, row.names = FALSE)

Output = data.frame(Species = sort(edf$Species))
iter = 1
for (i in newdata) {
  current = i
  iter = iter + 1
  MEANS <- ddply(i, .(Species), summarize, mean=mean(scapula.6))
  Output[,iter] = MEANS$mean
}
Filename = "Data/MarsMI_Scapula_AgnelCaudalBorder.csv"
write.csv(Output, Filename, row.names = FALSE)

Output = data.frame(Species = sort(edf$Species))
iter = 1
for (i in newdata) {
  current = i
  iter = iter + 1
  MEANS <- ddply(i, .(Species), summarize, mean=mean(humerus))
  Output[,iter] = MEANS$mean
}
Filename = "Data/MarsMI_Humerus_HeadLengthSaggital.csv"
write.csv(Output, Filename, row.names = FALSE)


Output = data.frame(Species = sort(edf$Species))
iter = 1
for (i in newdata) {
  current = i
  iter = iter + 1
  MEANS <- ddply(i, .(Species), summarize, mean=mean(humerus.1))
  Output[,iter] = MEANS$mean
}
Filename = "Data/MarsMI_Humerus_HeadLengthFrontal.csv"
write.csv(Output, Filename, row.names = FALSE)

Output = data.frame(Species = sort(edf$Species))
iter = 1
for (i in newdata) {
  current = i
  iter = iter + 1
  MEANS <- ddply(i, .(Species), summarize, mean=mean(humerus.2))
  Output[,iter] = MEANS$mean
}
Filename = "Data/MarsMI_Humerus_olecranonL.csv"
write.csv(Output, Filename, row.names = FALSE)

Output = data.frame(Species = sort(edf$Species))
iter = 1
for (i in newdata) {
  current = i
  iter = iter + 1
  MEANS <- ddply(i, .(Species), summarize, mean=mean(humerus.3))
  Output[,iter] = MEANS$mean
}
Filename = "Data/MarsMI_Humerus_trocheaL.csv"
write.csv(Output, Filename, row.names = FALSE)

Output = data.frame(Species = sort(edf$Species))
iter = 1
for (i in newdata) {
  current = i
  iter = iter + 1
  MEANS <- ddply(i, .(Species), summarize, mean=mean(humerus.4))
  Output[,iter] = MEANS$mean
}
Filename = "Data/MarsMI_Humerus_AngleTroch.csv"
write.csv(Output, Filename, row.names = FALSE)


Output = data.frame(Species = sort(edf$Species))
iter = 1
for (i in newdata) {
  current = i
  iter = iter + 1
  MEANS <- ddply(i, .(Species), summarize, mean=mean(femur))
  Output[,iter] = MEANS$mean
}
Filename = "Data/MarsMI_Femur_trochantericfossaLength.csv"
write.csv(Output, Filename, row.names = FALSE)


Output = data.frame(Species = sort(edf$Species))
iter = 1
for (i in newdata) {
  current = i
  iter = iter + 1
  MEANS <- ddply(i, .(Species), summarize, mean=mean(femur.1))
  Output[,iter] = MEANS$mean
}
Filename = "Data/MarsMI_Femur_trochantericfossaWidth.csv"
write.csv(Output, Filename, row.names = FALSE)

Output = data.frame(Species = sort(edf$Species))
iter = 1
for (i in newdata) {
  current = i
  iter = iter + 1
  MEANS <- ddply(i, .(Species), summarize, mean=mean(femur.2))
  Output[,iter] = MEANS$mean
}
Filename = "Data/MarsMI_Femur_circonference.csv"
write.csv(Output, Filename, row.names = FALSE)


Output = data.frame(Species = sort(edf$Species))
iter = 1
for (i in newdata) {
  current = i
  iter = iter + 1
  MEANS <- ddply(i, .(Species), summarize, mean=mean(femur.3))
  Output[,iter] = MEANS$mean
}
Filename = "Data/MarsMI_Femur_FemurLength.csv"
write.csv(Output, Filename, row.names = FALSE)



Output = data.frame(Species = sort(edf$Species))
iter = 1
for (i in newdata) {
  current = i
  iter = iter + 1
  MEANS <- ddply(i, .(Species), summarize, mean=mean(Tibia))
  Output[,iter] = MEANS$mean
}
Filename = "Data/MarsMI_Tibia_TibiaLength.csv"
write.csv(Output, Filename, row.names = FALSE)

Output = data.frame(Species = sort(edf$Species))
iter = 1
for (i in newdata) {
  current = i
  iter = iter + 1
  MEANS <- ddply(i, .(Species), summarize, mean=mean(Calcaneus))
  Output[,iter] = MEANS$mean
}
Filename = "Data/MarsMI_Calcaneus_MaxLength.csv"
write.csv(Output, Filename, row.names = FALSE)

Output = data.frame(Species = sort(edf$Species))
iter = 1
for (i in newdata) {
  current = i
  iter = iter + 1
  MEANS <- ddply(i, .(Species), summarize, mean=mean(Calcaneus.1))
  Output[,iter] = MEANS$mean
}
Filename = "Data/MarsMI_Calcaneus_HeadWidth.csv"
write.csv(Output, Filename, row.names = FALSE)

Output = data.frame(Species = sort(edf$Species))
iter = 1
for (i in newdata) {
  current = i
  iter = iter + 1
  MEANS <- ddply(i, .(Species), summarize, mean=mean(Calcaneus.2))
  Output[,iter] = MEANS$mean
}
Filename = "Data/MarsMI_Calcaneus_HeadMaxWidth.csv"
write.csv(Output, Filename, row.names = FALSE)

Output = data.frame(Species = sort(edf$Species))
iter = 1
for (i in newdata) {
  current = i
  iter = iter + 1
  MEANS <- ddply(i, .(Species), summarize, mean=mean(Calcaneus.3))
  Output[,iter] = MEANS$mean
}
Filename = "Data/MarsMI_Calcaneus_TuberWidth.csv"
write.csv(Output, Filename, row.names = FALSE)

Output = data.frame(Species = sort(edf$Species))
iter = 1
for (i in newdata) {
  current = i
  iter = iter + 1
  MEANS <- ddply(i, .(Species), summarize, mean=mean(Calcaneus.4))
  Output[,iter] = MEANS$mean
}
Filename = "Data/MarsMI_Calcaneus_TuberLength.csv"
write.csv(Output, Filename, row.names = FALSE)




```

```{r}


newdata = split(imp3.dat, f = imp3.dat$.imp)
Output = data.frame(Species = sort(edf$Species))
Filename = "MI_Data/MarsMI2lonly_Imp"
iter = 0
variables = colnames(Marsimp)
for (i in newdata) {
  current = i
  var_idx = which(colnames(i) %in% variables)
  iter = iter + 1
  Output = current[,var_idx]
  newfile = paste(Filename, iter, sep = "_")
  newfile = paste(newfile, "csv", sep = ".")
  write.csv(Output, newfile, row.names = FALSE)
}

```